A Heuristic Approach towards Drawings of Graphs with High Crossing Resolution

### 2018-05-23 paper/paper.tex

In this paper, we approach the crossing resolution maximization problem from a different perspective. We suggest a simple and quite intuitive randomization method for computing drawings with high crossing resolution, which, in a sense, mimics the way a human would try to increase the crossing resolution of a drawing. How would one try to increase the crossing resolution of a given drawing? First, she would try to identify the pair of edges that define the crossing resolution of the drawing (we call them \emph{critical} edges); then, she would try to move an endvertex of this pair (which we choose at random), hoping that by this move the crossing resolution will increase. Of course, we cannot consider all possible positions for the vertex to be moved. Therefore, we discretize them by considering a number of rays around the chosen vertex and by choosing (at random again) a single point at each of them.

@@

\begin{itemize}
%\item As long as the algorithm detects that within the last $\zeta$ iterations the crossing resolution of the graph has being improved, the vertex-pool contains only critical vertices. 
\item If during the last $\zeta$ iterations the crossing resolution has not been improved, then the vertex-pool becomes \emph{wider} containing all the vertices of the graph, and the algorithm is executed with this vertex-pool for $\zeta'$ iterations. 
\item After these $\zeta'$ iterations, the vertex-pool switches back to the critical vertices.
\end{itemize}
%
From our experimental evaluation, we observed that, while this approach turned to be quite effective for medium-size graphs, for larger and denser graph, unfortunately, it was not so efficient. What we observed is that in larger graphs it was unlike to change the embedding of the graph in a beneficial way for the algorithm to proceed, mainly because of the random choice of the vertices to be moved; in most iterations with the wider vertex-pool, the vertices chosen to be moved were not at the part of the drawing where the critical vertices resided.

@@

The termination condition of our algorithm is simple and depends also on a input parameter $\tau$, such that $\tau \geq \zeta + \zeta'$. More precisely, if the crossing resolution during the last $\tau$ iterations has not been improved, then we assume that the algorithm has converged and we do not proceed any more.

@@

The algorithm, as it has been described so far, does not necessarily produce grid drawings, i.e., drawings in which the vertices are at integer coordinates. However, it can be easily adjusted to produced such drawings. Recall that, at the $i$-th iteration step, our algorithm chooses uniformly at random a vertex $v_i$ from the vertex-pool and determines a set of $\rho$ candidate points $\pi_0,\ldots,\pi_{\rho-1}$. If we round these points to their closest grid points, say $\overline{\pi}_0,\ldots,\overline{\pi}_{\rho-1}$, and use them as candidates for the next position of $v_i$, then the obtained drawing will be grid, assuming that the starting drawing is grid (the latter requirement is not difficult to be achieved, e.g., by slightly adjusting the positions of the vertices of a circular layout, whose underlying circle has large radius, to be at grid points).

An alternative approach is to let our algorithm compute a (non-grid) drawing and then apply some technique to convert it to grid. Critical here is not to affect the crossing resolution too much during the conversion. Of course, if the crossing resolution of the obtained grid drawing is at least as large as the crossing resolution of the non-grid starting drawing, then we considered the computed grid drawing \emph{acceptable}. In our approach, we also accepted drawings whose crossing resolution is at least as large as the crossing resolution of the non-grid starting drawing minus an input parameter $\epsilon$, which we used as \emph{slack} for tolerating slightly worse solutions. For converting a non-grid drawing to grid, we repeated the following procedure, that is also based on randomization, until we compute an acceptable grid drawing: For each vertex of the non-grid starting drawing, chose uniformly at random one of the grid points from the so-called \emph{neighbourhood} of the vertex, and move the vertex to this point (assuming of course that there is no other vertex at this point), where the neighbourhood of a vertex was initially set to its four closest grid points. If after a number of iterations an acceptable grid drawing without vertex-vertex overlaps could not be reported, then we augmented the neighbourhood of all vertices to contain more grid points.

>>>>

In this paper, we approach the crossing resolution maximization problem from a different perspective. We suggest a simple and quite intuitive randomization method for computing drawings with high crossing resolution, which, in a sense, mimics the way a human would try to increase the crossing resolution of a drawing. How would one try to increase the crossing resolution of a given drawing? First, she would try to identify the pair of edges that define the crossing resolution of the drawing (we call them \emph{critical} edges); then, she would try to move an endvertex of this pair (which we choose at random), hoping that by this move the crossing resolution will increase. Of course, we cannot consider all possible positions for the vertex to be moved. Instead, we consider a small set of randomly generated ones. Then, if there exists a position among them, that does not lead to a reduction of the crossing resolution, we simply move the vertex to this~position.

@@

\begin{itemize}
%\item As long as the algorithm detects that within the last $\zeta$ iterations the crossing resolution of the graph has being improved, the vertex-pool contains only critical vertices. 
\item If during the last $\zeta$ iterations the crossing resolution has not been improved, then the vertex-pool becomes \emph{wider} containing all the vertices of the graph, and the algorithm is executed with this vertex-pool for $\zeta'$ iterations. 
\item After these $\zeta'$ iterations, the vertex-pool switches back to the critical vertices.
\end{itemize}
%
From our extensive practical analysis, we quickly realized that, while this approach turned out to be quite effective for medium-size graphs, for larger and denser graph, unfortunately, it was not so efficient. What we observed is that in larger graphs it was unlikely to change the embedding of the graph in a beneficial way for the algorithm to proceed, mainly because of the random choice of the vertices to be moved; in most iterations with the wider vertex-pool, the vertices chosen to be moved were not in the part of the drawing where the critical vertices resided.

@@

The termination condition of our algorithm is simple and depends also on an input parameter $\tau$, such that $\tau \geq \zeta + \zeta'$. More precisely, if the crossing resolution during the last $\tau$ iterations has not been improved, then we assume that the algorithm has converged and we do not proceed any more.

@@

The algorithm, as it has been described so far, does not necessarily produce grid drawings, i.e., drawings in which the vertices are at integer coordinates. However, it can be easily adjusted to produce such drawings. Recall that, at the $i$-th iteration step, our algorithm chooses uniformly at random a vertex $v_i$ from the vertex-pool and determines a set of $\rho$ candidate points $\pi_0,\ldots,\pi_{\rho-1}$. If we round these points to their closest grid points, say $\overline{\pi}_0,\ldots,\overline{\pi}_{\rho-1}$, and use them as candidates for the next position of $v_i$, then the obtained drawing will be grid, assuming that the starting drawing is grid (the latter requirement is not difficult to be achieved, e.g., by slightly adjusting the positions of the vertices of a circular layout, whose underlying circle has large radius, to be at grid points).

An alternative approach is to let our algorithm compute a (non-grid) drawing and then apply some technique to convert it to grid. Critical here is not to affect the crossing resolution too much during the conversion. Of course, if the crossing resolution of the obtained grid drawing is at least as large as the crossing resolution of the non-grid starting drawing, then we considered the computed grid drawing \emph{acceptable}. In our approach, we also accepted drawings whose crossing resolution is at least as large as the crossing resolution of the non-grid starting drawing minus an input parameter $\epsilon$, which we used as \emph{slack} for tolerating slightly worse solutions. For converting a non-grid drawing to grid, we repeated the following procedure, that is also based on randomization, until we compute an acceptable grid drawing: For each vertex of the non-grid starting drawing, chose uniformly at random one of the grid points from the so-called \emph{neighbourhood} of the vertex, and move the vertex to this point (assuming of course that there is no other vertex at this point), where the neighbourhood of a vertex was initially set to its four closest grid points. If after a number of iterations an acceptable grid drawing without vertex-vertex overlaps could not be reported, then we augmented the neighbourhood of all vertices to contain more grid points, in particular, the grid points that are in horizontal and vertical distance~$1$ from the grid points of the current neighborhood.
